Interpreting Random Forest Predictions for Firearm Identification Using LIME

### 2018-11-06 paper/paper.Rmd
worked more on LIME description

LIME was developed by Marco Tulio Ribeiro, Sameer Singh, and Carlos Guestrin in response to the issue of black box prediction models, which are complex models that may produce accurate predictions but do not allow for interpretability. The LIME algorithm can be applied to any predictive model. It provides an explanation for an individual prediction of interest by fitting a linear regression model in a local region around the prediction. Linear regression models contain a well known form and coefficients that can be interpreted.

The developers of LIME wrote a Python package to implement LIME, and Thomas Lin Pedersen adapted their work to create an R package called `lime`. This paper will use version `r packageVersion("lime")` of the R package `lime`. The steps below explain the procedure that the R package is using to apply the LIME algorithm to the bullet matching predictions on the Hamby 224 clone dataset made by the random forest model from Hare. For simplicity, the steps are described as what happens to one case in the test data. Thus, the steps (2) through (7) are repeated for each observation in the testing dataset.

@@

```{r echo = FALSE}
# Apply the lime function if the file does not already exist
if(!file.exists("../data/hamby224_lime.rds")) {

@@

```{r echo = FALSE}
# Function to use for creating bin labels in the test_explain dataset
bin_labeller <- function(feature, feature_value){

@@

    # Create a dataframe of bins cuts
    bin_cuts <- data.frame(matrix(unlist(hamby224_lime$bin_cuts), nrow = 9, byrow = TRUE)) %>%
      rename("Q0" = X1, "Q25" = X2, "Q50" = X3, "Q75" = X4, "Q100" = X5) %>%
      mutate(Feature = rf_features) %>%
      select(Feature, Q0:Q100)

@@

```{r}
check <- hamby224_test_explain %>%
  select(-data, -prediction) %>%
  group_by(case) %>%
  slice(1) %>%
  select(case, set, rfscore, model_prediction) %>%
  mutate(diff = rfscore - model_prediction,
         mean = (rfscore + model_prediction) / 2)

@@

```{r fig.width = 6, fig.height = 4}
ggplot(check, aes(x = rfscore, y = model_prediction)) + 
  geom_point() + 
  facet_grid( ~ set) + 
  theme_bw() + 
  stat_smooth(method = "lm")
```

@@

```{r fig.width = 6, fig.height = 4}
ggplot(check, aes(x = rfscore, y = abs(diff))) + 
  geom_point() + 
  facet_grid( ~ set) +
  theme_bw()
```

@@

```{r fig.width = 6, fig.height = 4}
ggplot(check, aes(x = rfscore, y = diff)) + 
  geom_point() + 
  facet_grid( ~ set) +
  geom_hline(yintercept = 0, color = "blue") +
  theme_bw()
```

@@

```{r}
hamby224_test_explain %>%
      filter(set == "Set 1") %>%
      mutate(rfscore = round(rfscore, 3)) %>%
      select(case, bullet1, bullet2, land1, land2, rfscore) %>%
      distinct() %>%
      ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
                 text = paste('Bullets Compared: ', bullet1, "-", land1, 
                              "vs", bullet2, "-", land2,
                              '\nRandom Forest Score: ', 
                              ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
      geom_tile(aes(fill = rfscore)) +
      facet_grid(bullet2 ~ bullet1, scales = "free") +
      theme_minimal() +
      scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            strip.text.x = element_blank(),
            strip.text.y = element_blank()) +
      labs(x = "Known 1                            Known 2                            Questioned", 
           y = "Known 1                            Known 2                            Questioned", 
           fill = "RF Score")
```

>>>>

Marco Tulio Ribeiro, Sameer Singh, and Carlos Guestrin developed LIME in response to the interpretabiliby issue of many machine learning techniques. Many models used in machine learning are referred to as black box prediction models since they are complex models that may provide accurate predictions but are not able to be interpreted. That is, it is not possible to understand the relationship between the features and the response variable. Without being able to interpret the models, it becomes difficult to diagnose why a model is producing the predictions that it is. The developers claim that LIME allows users to check which variables are driving the predictions made by the model. This further allows users to determine whether their model is trustworthy.

LIME was designed to be a generalized algorithm that can be applied to any predictive model to return an "explanation" for an individual prediction of interest. This explanation is meant to convey which features from the model are driving the prediction created by the complex model. In the next section, I will provide a detailed description of how the LIME algorithm obtains the explanation when applied to a random forest model with numeric features. For now, general overview of the LIME algorithm will be explained.

Suppose that a complex model has been fit on a training dataset with $k$ features and a response variable that may either be numeric or categorical. This trained model will be applied to a testing dataset containing the $k$ features to obtain predictions. It is of interest to understand which variables played an important role in the predictions. LIME will be applied to one of the cases in the testing dataset to obtain an explanation. To create the explanation, LIME starts by approximating the distribution of each of the $k$ features in the training dataset. It then simulates a large dataframe from these distributions. The simulated data are used to perform feature selection to determine the $m$ most important features. Then a weighted ridge regression model is fit to standardized versions of the selected $m$ features from the simulated data. The model assigns higher weights to observations closer to the observed case of interest. A linear model is used since it has a well known form and coefficients that can be interpreted. The coefficients from the model are extracted and used to "explain" the importance of the variables. Features with a larger absolute value of their coefficient are implied to play a larger role in the creation of the prediction.

The developers of LIME wrote a Python package to implement LIME, and Thomas Lin Pedersen adapted their work to create an R package called `lime`. This paper will use version `r packageVersion("lime")` of the R package `lime` in the analysis of the bullet matching data. The steps below explain the procedure that the R package is using to apply the LIME algorithm to the bullet matching predictions on the Hamby 224 clone dataset made by the random forest model from Hare. For simplicity, the steps are described as what happens to one case in the test data. Thus, the steps (2) through (7) are repeated for each observation in the testing dataset.

Let
  $$Y_{jk} = \begin{cases} 
  1 & \mbox{ if bullets } j \mbox{ and } k \mbox{ were fired from the same gun barrel }\\
  0 & \mbox{otherwise}
  \end{cases}$$
be the response variable in the training dataset, and $X_1,...,X_9$ correspond to the nine features in the training dataset. Let $X'_1,...,X'_9$ be the

@@

```{r}
# Apply the lime function if the file does not already exist
if(!file.exists("../data/hamby224_lime.rds")) {

@@

```{r}
if (!file.exists("../data/hamby_bins.csv")) {

  # Create a dataframe of bins cuts
  hamby_bins <- data.frame(matrix(unlist(hamby224_lime$bin_cuts), 
                                  nrow = 9, byrow = TRUE)) %>%
      rename("Q0" = X1, "Q25" = X2, "Q50" = X3, "Q75" = X4, "Q100" = X5) %>%
      mutate(Feature = rf_features) %>%
      select(Feature, Q0:Q100)

  # Save the bin cuts
  write.csv(hamby_bins, "../data/hamby_bins.csv", row.names = FALSE)

  # Load in the bins csv
  hamby_bins <- read.csv("../data/hamby_bins.csv")

```{r}
# Function to use for creating bin labels in the test_explain dataset
bin_labeller <- function(feature, feature_value, bin_cuts = hamby_bins){

@@

@@

```{r}
check <- hamby224_test_explain %>%
  select(-data, -prediction) %>%
  group_by(case) %>%
  slice(1) %>%
  select(case, set, rfscore, model_prediction, model_r2) %>%
  mutate(diff = rfscore - model_prediction,
         mean = (rfscore + model_prediction) / 2)

@@

THe points are also colored by the $R^2$ value from the simple model. This plot suggests that the simple models with a prediction close to 0.5 have a better fit, but the $R^2$ values are still low. I am not sure what to make of this. However, it does not appear that the $R^2$ is related to the probability from the random forest.

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = model_prediction)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) + 
  theme_bw() + 
  stat_smooth(method = "lm")
```

@@

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = abs(diff))) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) +
  theme_bw()
```

@@

```{r fig.width = 8, fig.height = 4}
ggplot(check, aes(x = rfscore, y = diff)) + 
  geom_point(aes(color = model_r2)) + 
  facet_grid( ~ set) +
  geom_hline(yintercept = 0, color = "blue") +
  theme_bw()
```

@@

```{r}
hamby224_test_explain %>%
      filter(set == "Set 1") %>%
      mutate(rfscore = round(rfscore, 3)) %>%
      select(case, bullet1, bullet2, land1, land2, rfscore) %>%
      distinct() %>%
      ggplot(aes(x = land1, y = land2, label = bullet1, label2 = bullet2,
                 text = paste('Bullets Compared: ', bullet1, "-", land1, 
                              "vs", bullet2, "-", land2,
                              '\nRandom Forest Score: ', 
                              ifelse(is.na(rfscore), "Missing due to tank rash", rfscore)))) +
      geom_tile(aes(fill = rfscore)) +
      facet_grid(bullet2 ~ bullet1, scales = "free") +
      theme_minimal() +
      scale_fill_gradient2(low = "darkgrey", high = "darkorange", midpoint = 0.5) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(x = "", y = "", fill = "RF Score")
```
